%\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate,dvipsnames]{lipics-v2021}
\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate,dvipsnames,anonymous]{lipics-v2021}


\usepackage{todonotes}
\usepackage{xspace}
\usepackage{mathtools}
\usepackage{multicol}
\usepackage{stmaryrd}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{nicefrac}
\usepackage{xparse}
\usepackage{array}
\usepackage{makecell}

%\usepackage[appendix=append]{apxproof} % move to appendix
%\usepackage[appendix=inline]{apxproof} % leave in place
\usepackage[appendix=strip]{apxproof} % remove proofs
%\newtheoremrep{lemma}[theorem]{Lemma}
\newtheoremrep{apxtheorem}[theorem]{Theorem}
\newtheoremrep{apxlemma}[theorem]{Lemma}

\nolinenumbers

\bibliographystyle{plainurl}% the mandatory bibstyle

\title{Reachability in 1-VASS with data} %TODO Please add

%\titlerunning{Dummy short title} %TODO optional, please use if title is longer than one line

\author{S{\l}awomir Lasota}{University of Warsaw}{}{https://orcid.org/0000-0001-8674-4470}{Partially supported by ERC grant INFSYS, agreement no. 950398 and 
by the NCN grant 2024/55/B/ST6/01674.}
\author{Mathieu Lehaut}{University of Warsaw}{}{https://orcid.org/0000-0002-6205-0682}{Partially supported by the NCN grant 2021/41/B/ST6/00535.}
\author{Julie Parreaux}{Univ Rennes, Inria, CNRS, IRISA}{}{https://orcid.org/0009-0009-2744-780X}{Partially supported by the ANR project BisoUS (ANR-22-CE48-0012).}
\author{Rados{\l}aw Pi{\'o}rkowski}{University of Oxford}{}{https://orcid.org/0000-0002-9643-182X}{}

\authorrunning{S.~Lasota, M.~Lehaut, J.~Parreaux, R.~Pi{\'o}rkowski} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{S{\l}awomir Lasota, Mathieu Lehaut, Julie Parreaux, Rados{\l}aw Pi{\'o}rkowski} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

%\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003766.10003770</concept_id>
<concept_desc>Theory of computation~Automata over infinite objects</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003752.10003766.10003773.10003775</concept_id>
<concept_desc>Theory of computation~Quantitative automata</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003752.10003790.10002990</concept_id>
<concept_desc>Theory of computation~Logic and verification</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Automata over infinite objects}
\ccsdesc[500]{Theory of computation~Quantitative automata}
\ccsdesc[500]{Theory of computation~Logic and verification}

\keywords{Petri nets, vector addition systems, data vector addition systems, reachability problem} %TODO mandatory; please add comma-separated list of keywords

\category{} %optional, e.g. invited paper

%\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional
\relatedversion{}
\relatedversiondetails{Extended Version}{https://arxiv.org/abs/2601.04902}

%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

%\acknowledgements{I want to thank \dots}%optional

%\nolinenumbers %uncomment to disable line numbering


\input{macros}


\begin{document}

\maketitle

%TODO mandatory: add short abstract of the document
\begin{abstract}
We study the complexity of the reachabiltiy problem in data vector addition systems 
with states (data VASS)
in dimension 1.
\end{abstract}


\newpage

\section{The model}

\emph{Data Petri nets} (resp.~\emph{data \vass}) are like plain Petri nets 
(resp.~\vass) except that every 
token carries a value from a given data universe, and fireability of transitions is 
conditioned by constraints on the data values involved.
In principle, the data universe may be any fixed relational structure, however
complex data universes often lead to undecidability of verification problems.
We consider the simplest data universe, the equality universe, which 
is a countable infinite set $\atoms$ (whose elements are called \emph{atoms}) 
with just equality relation: $(\atoms, =)$. 
In this case, fireability of a transition is conditioned by equalities and non-equalities
of data values carried by the tokens involved in a transition.
This is the most often studied data universe, for various models of computation
\cite{}.

In this paper we study the reachability problem in \dvass in dimension 1 (\donevass). 

\para{Definition of \vass} \sla{TODO}

\para{Definition of \donevass}
By an (integer) \emph{data vector} we mean any function $\vr v$
from atoms to integers that has finite 
support, namely assigns non-zero to only finitely many atoms, written
$\vr v : \atoms \finto \Z$.
Thus the \emph{support} of $\vr v$ is the set $\setof{a\in\atoms}{\vr v(a)\neq 0}$.
We define the \emph{norm} of $\vr v$ as the largest absolute value apparing in $\vr v$,
and the \emph{width} of $\vr v$ and the size of its support,
denote as $\norm {\vr v}$ and $\width {\vr v}$, respectively.
As expected, addition of data vectors is pointwise.

A \donevass $\V = (Q, T_0)$ consists of a finite set $Q$ of control states, and 
a finite set of transitions
$T_0\finsubseteq Q\times (\atoms\finto\Z)\times Q$. 
The data vector $\vr v$ is called the \emph{effect} of a transition $(q, \vr v, q')$.
We inplicitely extend the set $T_0$ by applying all permutations of atoms
$\sigma \in \aut \atoms$, namely 
\[
T := \setof{(q, \sigma(\vr v), q')}{(q, \vr v, q')\in T_0, \ \sigma \in \aut \atoms}, 
\]
where $\sigma(\vr v)$ stands for the natural point-wise action of permutations of 
atoms on data vectors:
\[ \sigma(\vr v)(\sigma(a)) = \vr v(a). \]
The set $T$ is thus infinite but \emph{orbit-finite}.
\Mywlog we may assume that $T_0$ nonredundant, namely no transition in $T_0$ maps
by an atom permutation to another transition in $T_0$.
In this case each element of $T_0$ represents an \emph{orbit} of $T$. 

\begin{remark}
Due to the closure under permutations, one could equivalently define the
effect of a transition as a finite multiset of nonzero integers, thus specifying non-zero
entries of a data vector without assigning the numbers to specific atoms.
\end{remark}

A configuration of $\V$ is a pair $(q, \vr w) \in Q\times (\atoms\finto\N)$ consisting of
a control state $q$ and a \emph{nonnegative} data vector $\vr w$, written as $q(\vr w)$.
For every nonnegative data vector $\vr w$, a transition $(q, \vr v, q') \in T$ induces 
a \emph{step} from configuration
$q(\vr w)$ to configuration $q'(\vr w+\vr v)$ of $\V$ as long as $\vr v + \vr w$ 
is still nonnegative, written
\[
q(\vr w) \step q'(\vr w + \vr v),
\]
exactly as in plain \vass or \onevass.
The \emph{reachability problem} asks, given a \donevass and two configurations, source $s$
and target $t$, whether there is a sequence of steps (called a \emph{run} later on) 
from $s$ to $t$, that is, whether $s \step^* t$.

As usual, we will distinguish between binary and unary encoding of numbers appearing in
the input.
%
For measuring the size of instance, we will use 
the \emph{norm} and \emph{width} of $\V$, denoted $\norm{\V}$ and $\width{\V}$, resp.,
and defined aa the largest norm, resp. the largest width, of an effect
in $T_0$.
We also naturally extend these measures to reachability instances $(\V,s,t)$ by taking
the vectors of $s$ and $t$ additionally into account.

As in plain (data-less) models, we have a tight connection between \vass, \vas and Petri
nets (\pn).
% Dimension of \vass (or \vas) corresponds to the number of places of Petri nets.
A slight restriction of the model of \donevass is \donevas, the stateless fragment.
Then, 
a slight generalisation of \donevas is data Petri nets with one place,
\emph{\donepn}.
The reachability problem in \donepn is known to be decidable \cite{}.

\begin{example}
\sla{TODO: an example of data 1-VASS or 1-VAS, maybe a picture}
\end{example}

\para{Symmetric \vass}
\donevass are like plain \vass of infinite dimension which are \emph{symmetric}, namely
whose transitions are invariant under all permutations of dimensions.
In this context it is natural to investigate symmetric \vass of finite dimension,
a synactic fragment of \vass.
The reachability problem in symmetric \vass, define along the same lines as for
\donevass, is know to be decidable in \pspace \cite{KL25},
even under the binary encoding, and under succinct representation that assumes that
only the representatives $T_0$ of orbits of $T$ are given in the input.
The \pspace upper bound appplies to any fixed dimension, as well as to the more general
case when dimension of a symmetric \vass is a part of input.

Symmetric \vass and \donevass are tightly connected, namely
a \donevass $\V$ together with a number $d\geq\width{\V}$ naturally induces 
a symmetric $d$-\vass $\findim {\V} d$, as restriction of (configurations of)
$\V$ to an arbitrary subset of atoms of size $d$.


\section{\Decomp \trace{s}}

Any sequence of nonzero integers $h = n_1, \ldots, n_\ell$ such that every 
prefix sum is nonnegative, namely $n_1 + \ldots + n_i \geq 0$ for every $i$,
and the total sum is 0, namely $n_1 + \ldots + n_\ell = 0$, we call a
\emph{\trace}.
A \trace is \emph{\decomp} if it is an interleaving of two shorter traces,
otherwise it is \emph{\indecomp}.
The norm $\norm{h}$ of a \trace $h$ is the largest among absolute values of all $n_i$. 

\begin{example}
\sla{TODO}
\end{example}

As our first combinatorial tool we prove that every \trace of length exceeding a polynomial
of its norm is \decomp:
\begin{lemma} \label{lem:decomp}
    There is a polynomial $p$ such that every \trace $h$ of length larger than $p(\norm h)$
    is \decomp.
\end{lemma}
%
\begin{proof}

    The polynomial $p$ will be revealed in the proof below.

    Consider a \trace $h = n_1 \ldots n_\ell$ and let $N$ be its maximal prefix-sum.
    \para{Case I: $N > $}
    \sla{hill cutting}

    \para{Case II: $N \leq $}
    \sla{GPT-sequences}

\end{proof}

\begin{corollary} \label{cor:decomp}
Every \trace $h$ is an interleaving of \trace{s} of length at most $p(\norm{h})$.
\end{corollary}


\section{Complexity of the reachability problem}

\begin{theorem} \label{thm:upper-bound}
The \donevass reachability problem is decidable in:
\begin{itemize} 
    \item \expspace under binary encoding;
    \item \pspace under unary encoding.
\end{itemize}
\end{theorem}
%
\begin{proof}
We design a reduction to the symmetric \vass rachability problem which is polynomial
un unary size of a \donevass reachability instance.
More specifically, we prove that 
every \donevass reachabiltity instance $(\V, s, t)$ is equivalent to
the symmetric $d$-\vass reachability instance $(\findim \V d, \overline s, \overline t)$ 
as long as 
the dimension $d$ is large enough, namely 
\[
d\ \geq \ d_0 \ := \ (p(\norm{\V}) + 2)\cdot\width{\V} + 1,
\]
for  the polynomial $p$ of \Cref{cor:decomp}.
% 
The equivalence means that $s \step^* t$ is $\V$ if and only if
$\overline s \step^* \overline t$ in $\findim {\V} d$.
The reduction essentially preserves the set $T_0$ of orbit-representatives of 
transitions, except that it pads every effect with sufficiently many zeros to get
vectors of required dimension $d$.
Likewise for source and target configurations $s$ and $t$, thus yielding their padded
versions $\overline s$ and $\overline t$.

The size of the symmetric $d$-\vass instance is thus polynomial in the size of 
$(\V,s,t)$ in case of
unary encoding, and exponential in the size of $\V$ in case of binary encoding.
The complexity claims of \Cref{thm:upper-bound} follow immediately.
\sla{Can we do better for binary encoding, 
namely can we solve symmetric $d$-\vass in \pspace even if the dimension
is exponential? The exponentially-long vectors would then be represented succinctly, 
by listing (positions, nonzero value) pairs.}

\medskip

Suppose $\V$ has a run from $s$ to $t$, and consider a fixed such run $\rho$.
For the correctness of the above reduction, it is enough to prove that
$\V$ has a run from $s$ to $t$ that uses at most $d_0$ atoms.
Formally, we are going to prove the following 
bound on the support-size of a run, where the support of a run is defined
as the union of supports of all vectors appearing in it:
\begin{claim} \label{claim:sup-bound}
$\V$ has a run from $s$ to $t$ whose support has size at most $d_0$.
\end{claim}

We need some preparation before embarking on the proof of the claim.

Let $S\finsubseteq \atoms$ be the support of the run $\rho$.
The set $S$ includes the supports of $s$ and $t$. The atoms in $S$ not belonging to
the two supports we call \emph{ordinary}. 
The run therefore starts and ends with 0 assigned to all ordinary atoms. 
When proving \Cref{claim:sup-bound} we will bound the largest number of
ordinary atoms appearing in a configuration along the run.

For every ordinary atom in $a\in S$ we can consider 
the subsequence $\rho_a$ of $\rho$ consisting 
of steps that uses atom $a$ (i.e., those steps, where the support of the effect
contains atom $a$). 
This subsequence represents thus the complete history of atom $a$ along the whole run. 
We define the \emph{\trace} $h_a$ of $a$ as the sequence 
of effects of transitions from $\rho_a$ on $a$.
This is a sequence of nonzero integers that is indeed a \trace:
every prefix sum is nonnegative, and the total sum is zero.

\medskip

We prove \Cref{claim:sup-bound} by transforming $\rho$ into a run of support size
at most $d_0$, in two steps.
As the first step, using \Cref{cor:decomp} (note that $\norm{h_a}\leq\norm{\V}$), 
for every ordinary atom $a$ we split $h_a$ into an interleaving of traces
$h_1, \ldots, h_m$, each of length at most $f(\norm{\V})$.
In every such subsequence, separately, we rename $a$ to a \emph{fresh} atom, i.e.,
an atom outside of $S$. 
In this way we get a new run $\rho'$ from $s$ to $t$ satisfying:

\begin{claim} \label{claim:trace-bound}
% $\V$ has a run $\rho'$ from $s$ to $t$ where 
The trace of every ordinary atom in $\rho'$
has length at most $p(\norm{\V})$.
\end{claim}
%
The new run may thus use way more atoms that $\rho$, but each ordinary atom
is used by a polynomially bounded number of step.

As the second step, we merge ordinary atoms in order to decrease their number below the
required threshold.
To this aim we define a graph whose nodes are all ordinary atoms apearing in $\rho'$,
and an edge connects atoms $a$ and $b$ if some step in $\rho'$ uses both $a$ and $b$.
Due to the bound on the lengths of traces of oedinary atoms given in \Cref{claim:trace-bound},
the degree of the graph is bounded by $B:=p(\norm{\V})\cdot \width{\V}$, 
and therefore the nodes of the graph can be colored using $B+1$ colors so that 
no edge connects two nodes of the same color.
We merge all ordinary atoms whose nodes are colored by the same color (for instance,
by remaming all of them to a chosen one). 
In this way we obtain a run $\rho''$ from $s$ to $t$ with at most $B+1$ ordinary atoms. 
Bounding the support sizes of $s$ and $t$ by $\width{\V}$, we deduce that
the support of $\rho''$ has size at most
\[
B+1 + 2\cdot\width{\V} \ = \  (p(\norm{\V}) + 2)\cdot\width{\V}+1 \ = \ d_0.
\]
\Cref{claim:sup-bound} is thus proved, and hence so is \Cref{thm:upper-bound}.


\end{proof}

Our lower bounds apply to stateless fragment:

\begin{theorem}
The \donevas reachability problem is:
\begin{itemize}
\item \pspace-hard under binary encoding;
\item \np-hard under unary encoding.
\end{itemize}
\end{theorem}
%
\begin{proof}
\end{proof}

\begin{remark}
The proof of \np lower bound applies to symmetric \vas reachability problem,
assuming the dimension is not fixed but given as a part of input.
\end{remark}

\newpage


%%
%% Bibliography
%%

%% Please use bibtex,
\bibliography{bib.bib}


\end{document}
